A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-06-23 paper/paper.tex

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} graph or \emph{RAC} graph, for short; see Figure~\ref{fig:examples}. For these graphs, several results, mostly of theoretical nature, are known (refer to Section~\ref{sec:relatedwork} for a short overview). Notably, RAC graph are sparse (they contain at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, where $n$ denotes the number of  vertices), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

@@

From a practical point of view, we are only aware of two methods that aim at drawings with high crossing resolution; both of them are adjustments of force-directed algorithms~\cite{DBLP:journals/congnum/Eades84}. The first one is due to Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, while the~second one is due to Argyriou et al.\ Common in both algorithms is that they apply appropriate forces on the endvertices of every pair of crossing edges. Each of them uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea of both is the same: the smaller the crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

@@

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works.
%In Section~\ref{sec:preliminaries}, we introduce preliminary notions and the notation used throughout the paper.
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.\ in Section~\ref{sec:experiments}, where we also discuss our insights from this project. In the appendix, we provide more experimental results on grid restricted drawings, on more test sets and on the graphs from the graph drawing competition in 2017.

@@

An immediate observation emerging from the above overview is that the~focus has been primarily on theoretical aspects of the problem. Most of the approaches that could be useful in practice are based on force-directed techniques~\cite{DBLP:books/ph/BattistaETT99,DBLP:journals/congnum/Eades84}. COWA is a system that supports conceptual web site traffic analysis~\cite{DBLP:conf/apvis/DidimoLR10}; its algorithmic core is a force-directed heuristic to compute simultaneous embeddings of two non-planar graphs with high crossing resolution.
%
Didimo et al.~\cite{DBLP:conf/gd/DidimoLR10} describe topology-driven force-directed heuristics to achieve good trade-offs in terms of number of edge crossings, crossing resolution, and geodesic edge tendency; the obtained drawings, however, are not straight-line.
%
For straight-line drawings, Nguyen et al.~\cite{DBLP:conf/gd/NguyenEHH10} suggest a quadratic-program to increase the crossing angles of circular drawings.
%
Of more general scope are the already mentioned force-directed algorithms of Argyriou et al.\ and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

@@

%With this procedure, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm.
What we quickly realized from our practical analysis, is that the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing only at the critical vertices, it is~highly possible that the algorithm will get trapped to some local maxima after a number of iterations. So, special care is needed to avoid these bottlenecks, especially when the input graph is large. We discuss ways to avoid them later in this section.

@@

Let $v_i$ be the vertex of $\Gamma_{i-1}$ that has been chosen to be moved at the $i$-th iteration.
%Recall that the crossing resolution $c(\Gamma_{i})$ of drawing $\Gamma_{i}$ that is obtained after the $i$-th iteration of the algorithm must be at least as large as the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$ (by the invariant of our algorithm), that is, $c(\Gamma_i) \ge c(\Gamma_{i-1})$.
To compute the position of $v_i$ in the next drawing $\Gamma_i$, we consider a set of $\rho$ rays $r_0,r_1,\ldots,r_{\rho-1}$ that all emanate from $p(v_i)$ in $\Gamma_{i-1}$, such that the angle formed by ray $r_j$, with $j=0,1,\ldots,\rho-1$, and the horizontal axis equals to $2j\pi/\rho$, where $\rho>0$ is an integer parameter of the algorithm. These rays are then rotated by an angle that is chosen uniformly at random in the interval $[0,2\pi]$; see Fig.~\ref{fig:algo}. The position of vertex $v_i$ in $\Gamma_i$ will eventually be along one of the rays $r_0,r_1,\ldots,r_{\rho-1}$. More precisely, for each ray $r_i$ we chose a distance value $\delta_i$ uniformly at random from the interval $[\delta_{min},\delta_{max}]$, where $\delta_{min}$ and $\delta_{max}$ are two positive parameters of the algorithm. For each $j=0,1,\ldots,\rho-1$, a new point $\pi_j$ is obtained by translating $p(u)$ along $r_j$ by a distance $\delta_j$; point $\pi_j$ is \emph{feasible}, if the crossing resolution of the drawing obtained by placing vertex $v_i$ at $\pi_j$ and by keeping all other vertices of $G$ in their positions in $\Gamma_{i-1}$ is at least as large as the crossing resolution of $\Gamma_{i-1}$, and there is no vertex of $\Gamma_{i-1}$ at $\pi_j$.

@@

If none of the points $\pi_j$, with $j=0,1,\ldots,\rho-1$ is feasible, then the position of $v_i$ in $\Gamma_i$ is $p(v_i)$, i.e., same as in $\Gamma_{i-1}$, since $c(\Gamma_i) \geq c(\Gamma_{i-1})$ must hold. If there is one or more feasible points, then one may consider two different approaches to determine the position of $v_i$ in $\Gamma_i$. The most natural is to chose the feasible point that maximizes the crossing resolution of the obtained drawing. As an alternative, one may rely again on randomization and chose uniformly at random one of the feasible points as the position of $v_i$ in $\Gamma_i$. We note that we did not observe any significant difference between these two approaches (in terms of the crossing resolution of the obtained drawings), so we simply adopted the first one.

@@

Our second approach is based on parameters $\rho$, $\delta_{min}$ and $\delta_{max}$ of the algorithm. Our idea was that if the algorithm gets trapped to a locally optimal solution, then a ``drastic'' or ``sharp'' move may help to escape. We turned this idea into an algorithmic implementation as follows:
%
\begin{inparaenum}[(i)]
\item if during the last $\zeta$ iterations the crossing resolution has not been improved, we double the values of $\rho$, $\delta_{min}$ and $\delta_{max}$, and the algorithm is executed with these values for $\zeta'$ iterations;
\item afterwards, $\rho$, $\delta_{min}$ and $\delta_{max}$ switch back to this initial value.
\end{inparaenum}
%
Of course, this approach may lead to drawings with larger area, but this is expected, as it turns out that drawings with high crossing resolution may require large area~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/tcs/BrandenburgDEKL16}.

@@

If the algorithm had to compute all crossing points and the corresponding angles for each candidate position of each iteration, then it would not be useful in practice. Instead, we adopted a different approach, which turned out to be quite efficient in practice. Recall that we denoted by $v_i$ the vertex chosen at the $i$-th iteration step, and by $\pi_0,\ldots,\pi_{\rho-1}$ the candidate points to move $v_i$. Let $e_0,\ldots,e_{d_i-1}$ be the edges incident to $v_i$, where $d_i=deg(v_i)$. Next, for each edge $e_k$ with $k=0,\ldots,d_i-1$  we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges in $\Gamma_{i-1}$. Let $\phi_i$ be the minimum crossing angle computed; this is our reference angle. Also, for each candidate position $\pi_j$ with $j=0,\ldots,\rho-1$, and for each edge $e_k$ with $k=0,\ldots,d_i-1$, we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges of the drawing, assuming that $v_i$ is at $\pi_j$. Let $\chi_j$ be the minimum crossing angle computed with this approach, when $v_i$ is at position $\pi_j$. Clearly, $\pi_j$ is feasible only if $\chi_j \geq \phi_i$. Note that the complexity of this approach is $O(deg(v_i)m) = O(nm)$.

@@

% ==================================================================
\myparagraph{Total resolution}
% ==================================================================
%
%The notion of the total resolution of a drawing was introduced relatively recently with aim of ``balancing'' the measures of the crossing and of the angular resolution of a drawing~\cite{DBLP:journals/cj/ArgyriouBS13}. Formally, it is defined as the minimum of these two measures. It was not difficult to
Similarly as above, we could adjust our algorithm to yield drawings with high total resolution by simply taking into account also the angular resolution of the drawing. In particular, if the total resolution of the drawing is defined by its angular resolution, then the way we compute the critical vertices of this drawing has to change; the critical vertices must be the endvertices of the pairs of edges that define the angular resolution. Also, at each iteration of our algorithm we have to ensure that the total resolution does not decrease. We do so by rejecting candidate positions which yield a reduced total~resolution.

@@

A different behaviour can be observed in the number of iterations, which are required by the algorithms to converge; refer to Fig.~\ref{fig:cr-res-4}. We note here that we used different criteria to determine whether the algorithms of our experiment had converged. For our algorithms and for the force-directed algorithm by Huang et al., we assumed that the algorithm had converged, if the crossing resolution between 500 consecutive iterations was not improved by more than 0.001 degrees. For the algorithm by Argyriou et al., we decided to use a much more restricted convergence criterion, because the produced layouts can change vastly between  consecutive iterations. We made this choice mainly to have ``comparable'' number of iterations among the algorithms of the experiment. In this direction, we adopted the convergence criterion that the authors used in their previous experimental analysis\, that is, we assumed that the algorithm had converged, if the crossing resolution between two consecutive iterations was not improved by more than 0.001 degrees. Observe that even under this more restricted convergence criterion, the algorithm needs significantly more iterations to converge than the remaining three algorithms of the experiment; see Fig.~\ref{fig:cr-res-4}. The maximum number of iterations that each of the algorithms could perform in order to converge was set to 100.000, but that limit was never reached. We observe that both force-directed algorithms seem to require a great amount of iterations to converge for small graphs, where a drawing with really good crossing resolution is possible. However, for bigger graphs  the algorithm by Huang et al.\ requires the least amount of iterations. On the other hand, both the unrestricted and the restricted variant of our algorithm require comparable number of iterations to converge, but clearly more than the ones of the algorithm by Huang et al.

@@

% ==================================================================
\myparagraph{Discussion}
%
% ==================================================================
%
%In this paper, we introduced a new heuristic aiming to produce drawings of high crossing resolution, for which we also presented variants that take into account other common aesthetic criteria in Graph Drawing. Our experimental evaluation indicates that the new heuristic is competitive to state of the art force-directed algorithms, even when restricted to a given maximum aspect ratio. As a future direction, we plan to evaluate the performance of variants of our heuristic that compromise between even more aesthetic criteria.
%
While working on this project, we made some useful observations and  obtained some interesting insights. In particular, there is a recent hypothesis (also supported by experiments) that drawings in which the crossing angles are large are easy to read and understand. What we observed is that the drawings that are optimized only in terms of the crossing angles might be arbitrarily bad and may have several undesired properties. In particular, in these drawings it was very common to have adjacent edges to run almost in parallel and vertices to be very close to each other (nearly at the same spot). Hence, the angular resolution was very often poor, while the aspect ratio could be arbitrarily bad. The additional restrictions that we imposed regarding the angular resolution and the aspect ratio helped in significantly improving the readability of the drawings, without loosing too much of their quality in terms of the crossing resolution.

We conclude by noting that our motivation to work with this problem was our participation to GD2017 contest, where we performed miserably using a force-directed algorithm; for details see Appendix~\ref{app:contest}. As our evaluation shows, the performance of such algorithms is good, only when several aesthetic criteria are taken into account; our new approach is definitely more promising than our previous (as also evidenced by our experimental evaluation). The framework that we developed seems to be quite adaptable to optimize or to take into account also other desired aesthetic properties of a drawing (even although our initial plan was to optimize only the crossing resolution).

@@

Regarding the crossing resolution, we can observe that with increasing grid size, we could achieve better crossing resolution; see Fig.~\ref{fig:crossingResolutionGrid}. More precisely, a grid of size $10^2 \times 10^2$ was too restrictive for the vast majority of the graphs. As a result, the reported drawings were often the initial ones (as our algorithm could not improve them), especially for large graphs. Significantly fewer were the graphs for which our algorithm could not report an improved drawing, when grid size was set to $10^3 \times 10^3$.
%Starting from grid size $10^3 \times 10^3$, we could observe, for small graphs, the performance was close to the performance of the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}).
%For grid size $10^3 \times 10^3$, the performance declined for the larger graphs in the test set and we again could observe that for some of them the algorithm was not able to improve the initial layout.
For grid size $10^4 \times 10^4$, the drawings produced by our algorithm were on average by only $10^\circ$ worse than those produced by the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}), while the gap was closer for grid size $10^6 \times 10^6$.

The aspect ratio of the computed drawings was more or less the same regardless of the size of the underlying grid, with the exception of the drawings computed on the grid of size $10^2 \times 10^2$; see Fig.~\ref{fig:crossingResolutionAspectRatioGrid}. The fact that aspect ratio of these drawings was worse can also explained of course by the fact that in most cases an improved drawing could not be reported.

As expected, the smaller the underlying grid is, the more crossings the computed drawings contain; see Fig.~\ref{fig:crossingResolutionCrossingNumberGrid}. As a result, the unrestricted variant of our algorithm clearly outperforms all other ones. It is worth noting that the differences are clear between grid sizes $10^2 \times 10^2$, $10^3 \times 10^3$ and $10^4 \times 10^4$. Notably, there is only a slight improvement (in terms of the number of crossings) from grid size $10^4 \times 10^4$ to $10^6 \times 10^6$. %This can partially be explained with the different choice of the initial drawing.
%
On the other hand, the number of iterations needed for convergence increases with the grid size; see Fig.~\ref{fig:crossingResolutionTimeGrid}, with the exception of the grid of size $10^2 \times 10^2$, which clearly indicates our previous observation that for the vast majority of the graphs an improved drawing could not be reported.

@@

In this section, we report the results of our expertimental evaluation (on the crossing, total and angular resolutions) for the non-planar AT\&T graphs, which form a collection of 424 benchmark graphs (also known as Graph Catalog and North graphs; available at \texttt{http://graphdrawing.org/data}). Note that we did not impose any  grid constraints our algorithms. The corresponding results are illustrated in Figs.~\ref{fig:northCrossing}, \ref{fig:northTotal} and~\ref{fig:northAngular}. In general we observed that the variance of results is much larger than in the experiments on the Rome graphs. This manifests in spikes of large magnitude in the illustrations of the results and indicates that the structural properties of the graphs in this second test set varies vastly between different graph sizes.

@@

For the crossing resolution we observed that both variants of our algorithm again outperformed the two force-directed algorithms; see Fig.~\ref{fig:northCrossing-1}. Remarkable is the synchronous behaviour of all four algorithms regarding the crossing resolution, as the curves are nearly parallel. By all these results, we can classify the graphs into ``hard'' or ``easy'' when maximizing their crossing resolution. In particular, graphs with 50 to 70 vertices appear to be harder to improve than graphs with 70 to 80 vertices. Regarding the aspect ratio of the produced drawings, we observe that while our algorithms show a slight increase with the number of vertices, the behaviour for both force-directed algorithms appears to be unstable resulting in a large variance. Again the restricted variant of our algorithm and the two force directed approaches produce drawings with similar aspect ratio, which is much lower than the one of our unrestricted algorithm for larger graphs. All four algorithms behave nearly the same in terms of the number of crossings; see Fig.~\ref{fig:northCrossing-3}. In terms of the number of iterations, we observe that somewhat surprisingly the algorithm of Argyriou et al. converges in the least amount of iterations, while the remaining three algorithms behave nearly the same; see Fig.~\ref{fig:northCrossing-4}.

@@

Our primary motivation for this work was our participation to the graph drawing contest in 2017, where we miserably performed\footnote{\texttt{http://www.graphdrawing.de/contest2017/results.html}}. The topic was exactly the maximization of the crossing resolution. Our approach for the contest in 2017 was a mixture of the two force directed algorithms by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

>>>>

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} graph or \emph{RAC} graph, for short; see Figure~\ref{fig:examples}. For these graphs, several results, mostly of theoretical nature, are known (refer to Section~\ref{sec:relatedwork} for a short overview). Notably, RAC graphs are sparse (they contain at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, where $n$ denotes the number of  vertices), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

@@

From a practical point of view, we are only aware of two methods that aim at drawings with high crossing resolution; both of them are adjustments of force-directed algorithms~\cite{DBLP:journals/congnum/Eades84}. The first one is due to Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, while the~second one is due to Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}. Common in both algorithms is that they apply appropriate forces on the endvertices of every pair of crossing edges. Each of them uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea of both is the same: the smaller the crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

@@

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works.
%In Section~\ref{sec:preliminaries}, we introduce preliminary notions and the notation used throughout the paper.
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project. In the appendix, we provide experimental results on grid restricted drawings, on more test sets and on the graphs from the Graph Drawing Competition in 2017.

@@

An immediate observation emerging from the above overview is that the~focus has been primarily on theoretical aspects of the problem. Most of the approaches that could be useful in practice are based on force-directed techniques~\cite{DBLP:books/ph/BattistaETT99,DBLP:journals/congnum/Eades84}. COWA is a system that supports conceptual web site traffic analysis~\cite{DBLP:conf/apvis/DidimoLR10}; its algorithmic core is a force-directed heuristic to compute simultaneous embeddings of two non-planar graphs with high crossing resolution.
%
Didimo et al.~\cite{DBLP:conf/gd/DidimoLR10} describe topology-driven force-directed heuristics to achieve good trade-offs in terms of number of edge crossings, crossing resolution, and geodesic edge tendency; the obtained drawings, however, are not straight-line.
%
For straight-line drawings, Nguyen et al.~\cite{DBLP:conf/gd/NguyenEHH10} suggest a quadratic-program to increase the crossing angles of circular drawings.
%
Of more general scope are the already mentioned force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

@@

%With this procedure, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm.
What we quickly realized from our practical analysis, is that the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing only at the critical vertices, it is~highly possible that the algorithm will get trapped to some local maxima after a number of iterations. So, special care is needed to avoid these bottlenecks, especially when the input graph is large. We will discuss ways to avoid them later in this section.

@@

Let $v_i$ be the vertex of $\Gamma_{i-1}$ that has been chosen to be moved at the $i$-th iteration.
%Recall that the crossing resolution $c(\Gamma_{i})$ of drawing $\Gamma_{i}$ that is obtained after the $i$-th iteration of the algorithm must be at least as large as the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$ (by the invariant of our algorithm), that is, $c(\Gamma_i) \ge c(\Gamma_{i-1})$.
To compute the position of $v_i$ in the next drawing $\Gamma_i$, we consider a set of $\rho$ rays $r_0,r_1,\ldots,r_{\rho-1}$ that all emanate from $p(v_i)$ in $\Gamma_{i-1}$, such that the angle formed by ray $r_j$, with $j=0,1,\ldots,\rho-1$, and the horizontal axis equals to $2j\pi/\rho$, where $\rho>0$ is an integer parameter of the algorithm. These rays are then rotated by an angle that is chosen uniformly at random in the interval $[0,2\pi]$; see Fig.~\ref{fig:algo}. The position of vertex $v_i$ in $\Gamma_i$ will eventually be along one of the rays $r_0,r_1,\ldots,r_{\rho-1}$. More precisely, for each ray $r_i$ we choose a distance value $\delta_i$ uniformly at random from the interval $[\delta_{min},\delta_{max}]$, where $\delta_{min}$ and $\delta_{max}$ are two positive parameters of the algorithm. For each $j=0,1,\ldots,\rho-1$, a new point $\pi_j$ is obtained by translating $p(u)$ along $r_j$ by a distance $\delta_j$; point $\pi_j$ is \emph{feasible}, if the crossing resolution of the drawing obtained by placing vertex $v_i$ at $\pi_j$ and by keeping all other vertices of $G$ in their positions in $\Gamma_{i-1}$ is at least as large as the crossing resolution of $\Gamma_{i-1}$, and there is no vertex of $\Gamma_{i-1}$ at $\pi_j$.

@@

If none of the points $\pi_j$, with $j=0,1,\ldots,\rho-1$ is feasible, then the position of $v_i$ in $\Gamma_i$ is $p(v_i)$, i.e., same as in $\Gamma_{i-1}$, since $c(\Gamma_i) \geq c(\Gamma_{i-1})$ must hold. If there is one or more feasible points, then one may consider two different approaches to determine the position of $v_i$ in $\Gamma_i$. The most natural is to choose the feasible point that maximizes the crossing resolution of the obtained drawing. As an alternative, one may rely again on randomization and chose uniformly at random one of the feasible points as the position of $v_i$ in $\Gamma_i$. We note that we did not observe any significant difference between these two approaches (in terms of the crossing resolution of the obtained drawings), so we simply adopted the first one.

@@

Our second approach is based on parameters $\rho$, $\delta_{min}$ and $\delta_{max}$ of the algorithm. Our idea was that if the algorithm gets trapped to a locally optimal solution, then a ``drastic'' or ``sharp'' move may help to escape. We turned this idea into an algorithmic implementation as follows:
%
\begin{inparaenum}[(i)]
\item if during the last $\zeta$ iterations the crossing resolution has not been improved, we double the values of $\rho$, $\delta_{min}$ and $\delta_{max}$, and the algorithm is executed with these values for $\zeta'$ iterations;
\item afterwards, $\rho$, $\delta_{min}$ and $\delta_{max}$ switch back to this initial value.
\end{inparaenum}
%
This approach may lead to drawings with larger area, but this is ``expected'', as it turns out that drawings with high crossing resolution may require large area~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/tcs/BrandenburgDEKL16}.

@@

If the algorithm had to compute all crossing points and the corresponding angles for each candidate position of each iteration, then it would not be useful. Instead, we adopted a different approach, which turned out to be quite efficient in practice. Recall that we denoted by $v_i$ the vertex chosen at the $i$-th iteration step, and by $\pi_0,\ldots,\pi_{\rho-1}$ the candidate points to move $v_i$. Let $e_0,\ldots,e_{d_i-1}$ be the edges incident to $v_i$, where $d_i=deg(v_i)$. Next, for each edge $e_k$ with $k=0,\ldots,d_i-1$  we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges in $\Gamma_{i-1}$. Let $\phi_i$ be the minimum crossing angle computed; this is our reference angle. Also, for each candidate position $\pi_j$ with $j=0,\ldots,\rho-1$, and for each edge $e_k$ with $k=0,\ldots,d_i-1$, we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges of the drawing, assuming that $v_i$ is at $\pi_j$. Let $\chi_j$ be the minimum crossing angle computed with this approach, when $v_i$ is at position $\pi_j$. Clearly, $\pi_j$ is feasible only if $\chi_j \geq \phi_i$. Note that the complexity of this approach is $O(deg(v_i)m) = O(nm)$.

@@

% ==================================================================
\myparagraph{Total resolution}
% ==================================================================
%
%The notion of the total resolution of a drawing was introduced relatively recently with aim of ``balancing'' the measures of the crossing and of the angular resolution of a drawing~\cite{DBLP:journals/cj/ArgyriouBS13}. Formally, it is defined as the minimum of these two measures. It was not difficult to
Similarl as above, we could adjust our algorithm to yield drawings with high total resolution by simply taking into account also the angular resolution of the drawing. In particular, if the total resolution of the drawing is defined by its angular resolution, then the way we compute the critical vertices of this drawing has to change; the critical vertices must be the endvertices of the pairs of edges that define the angular resolution. Also, at each iteration of our algorithm we have to ensure that the total resolution does not decrease. We do so by rejecting candidate positions which yield a reduced total~resolution.

@@

A different behaviour can be observed in the number of iterations, which are required by the algorithms to converge; refer to Fig.~\ref{fig:cr-res-4}. We note here that we used different criteria to determine whether the algorithms of our experiment had converged. For our algorithms and for the force-directed algorithm by Huang et al., we assumed that the algorithm had converged, if the crossing resolution between 500 consecutive iterations was not improved by more than 0.001 degrees. For the algorithm by Argyriou et al., we decided to use a much more restricted convergence criterion, because the produced layouts can change vastly between  consecutive iterations. We made this choice mainly to have ``comparable'' number of iterations among the algorithms of the experiment. In this direction, we adopted the convergence criterion that the authors used in their previous experimental analysis\, that is, we assumed that the algorithm had converged, if the crossing resolution between two consecutive iterations was not improved by more than 0.001 degrees. Observe that even under this more restricted convergence criterion, the algorithm needs significantly more iterations to converge than the remaining three algorithms of the experiment; see Fig.~\ref{fig:cr-res-4}. The maximum number of iterations that each of the algorithms could perform in order to converge was set to 100.000, but that limit was never reached. We observe that both force-directed algorithms seem to require a great amount of iterations to converge for small graphs, where a drawing with really good crossing resolution is possible. However, for larger graphs  the algorithm by Huang et al.\ requires the least amount of iterations. On the other hand, both the unrestricted and the restricted variant of our algorithm require comparable number of iterations to converge, but clearly more than the ones of the algorithm by Huang et al.

@@

% ==================================================================
\myparagraph{Discussion}
%
% ==================================================================
%
%In this paper, we introduced a new heuristic aiming to produce drawings of high crossing resolution, for which we also presented variants that take into account other common aesthetic criteria in Graph Drawing. Our experimental evaluation indicates that the new heuristic is competitive to state of the art force-directed algorithms, even when restricted to a given maximum aspect ratio. As a future direction, we plan to evaluate the performance of variants of our heuristic that compromise between even more aesthetic criteria.
%
While working on this project, we made some useful observations and  obtained some interesting insights. In particular, there is a recent hypothesis (also supported by experiments) that drawings, in which the crossing angles, are large are easy to read and understand. What we observed is that the drawings that are optimized only in terms of the crossing angles might be arbitrarily bad and may have several undesired properties. In particular, in these drawings it was very common to have adjacent edges to run almost in parallel and vertices to be very close to each other (nearly at the same spot). Hence, the angular resolution and the aspect ratio were very often poor. The additional restrictions that we imposed regarding the angular resolution and the aspect ratio helped in significantly improving the readability of the drawings, without loosing too much of their quality in terms of the crossing resolution.

We conclude by noting that our motivation to work with this problem was our participation to GD2017 contest, where we performed miserably using a force-directed algorithm; for details see Appendix~\ref{app:contest}. As our evaluation shows, the performance of such algorithms is good, only when several aesthetic criteria are taken into account; our new approach is definitely more promising than our previous (as also evidenced by our experimental evaluation). The framework that we developed seems to be quite adaptable to optimize or to take into account also other desired aesthetic properties of a drawing (even though our initial plan was to optimize only the crossing resolution).

@@

Regarding the crossing resolution, we can observe that with increasing grid size, we could achieve better crossing resolution; see Fig.~\ref{fig:crossingResolutionGrid}. More precisely, a grid of size $10^2 \times 10^2$ was too restrictive for the vast majority of the graphs. As a result, the reported drawings were often the initial ones (as our algorithm could not improve them), especially for large graphs. Significantly fewer were the graphs for which our algorithm could not report an improved drawing, when the grid size was set to $10^3 \times 10^3$.
%Starting from grid size $10^3 \times 10^3$, we could observe, for small graphs, the performance was close to the performance of the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}).
%For grid size $10^3 \times 10^3$, the performance declined for the larger graphs in the test set and we again could observe that for some of them the algorithm was not able to improve the initial layout.
For grid size $10^4 \times 10^4$, the drawings produced by our algorithm were on average by only $10^\circ$ worse than those produced by the unrestricted version of our algorithm (double line in Fig.~\ref{fig:crossingResolutionGrid}), while the gap was closer for grid size $10^6 \times 10^6$.

The aspect ratio of the computed drawings was more or less the same regardless of the size of the underlying grid, with the exception of the drawings computed on the grid of size $10^2 \times 10^2$; see Fig.~\ref{fig:crossingResolutionAspectRatioGrid}. The fact that the aspect ratio of these drawings was worse can be explained of course by the fact that in most cases an improved drawing could not be reported.

As expected, the smaller the underlying grid is, the more crossings the computed drawings contain; see Fig.~\ref{fig:crossingResolutionCrossingNumberGrid}. As a result, the unrestricted variant of our algorithm clearly outperforms all other ones. It is worth noting that the differences are clear between grid sizes $10^2 \times 10^2$, $10^3 \times 10^3$ and $10^4 \times 10^4$. Notably, there is only a slight improvement (in terms of the number of crossings) from grid size $10^4 \times 10^4$ to $10^6 \times 10^6$. %This can partially be explained with the different choice of the initial drawing.
%
On the other hand, the number of iterations needed for convergence increases with the grid size (see Fig.~\ref{fig:crossingResolutionTimeGrid}), with the exception of the grid of size $10^2 \times 10^2$, which verifies our previous observation that for the vast majority of the graphs an improved drawing could not be reported.

@@

In this section, we report the results of our expertimental evaluation (on the crossing, total and angular resolutions) for the non-planar AT\&T graphs, which form a collection of 424 benchmark graphs (also known as Graph Catalog and North graphs; available at \texttt{http://graphdrawing.org/data}). Note that we did not impose any grid constraint on our algorithms. The corresponding results are illustrated in Figs.~\ref{fig:northCrossing}, \ref{fig:northTotal} and~\ref{fig:northAngular}. In general, we observed that the variance of the results is much larger than in the experiments on the Rome graphs. This manifests in spikes of large magnitude in the illustrations of the results and indicates that the structural properties of the graphs in this second test set varies vastly between different graph sizes.

@@

For the crossing resolution, we observed that both variants of our algorithm again outperformed the two force-directed algorithms; see Fig.~\ref{fig:northCrossing-1}. Remarkable is the synchronous behaviour of all four algorithms regarding the crossing resolution, as the curves are nearly parallel. By all these results, we can classify the graphs into ``hard'' or ``easy'' when maximizing their crossing resolution. In particular, graphs with 50 to 70 vertices appear to be harder to improve than graphs with 70 to 80 vertices. Regarding the aspect ratio of the produced drawings, we observe that while our algorithms show a slight increase with the number of vertices, the behaviour for both force-directed algorithms appears to be unstable resulting in a large variance. Again the restricted variant of our algorithm and the two force directed approaches produce drawings with similar aspect ratio, which is much lower than the one of our unrestricted algorithm for larger graphs. All four algorithms behave nearly the same in terms of the number of crossings; see Fig.~\ref{fig:northCrossing-3}. In terms of the number of iterations, we observe that somewhat surprisingly the algorithm of Argyriou et al. converges in the least amount of iterations, while the remaining three algorithms behave nearly the same; see Fig.~\ref{fig:northCrossing-4}.

@@

Our primary motivation for this work was our participation to the graph drawing contest in 2017, where we miserably performed\footnote{\texttt{http://www.graphdrawing.de/contest2017/results.html}}; the topic was the maximization of the crossing resolution. Our approach for the contest in 2017 was a mixture of the two force directed algorithms by Argyriou et al. and Huang et al.
